Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2024-12-24 _manuscript/_tex/index.tex
section 5 and 6

In the study of the health effects of outdoor air pollution, one area of
interest is the association between short-term, day-to-day changes in
particulate matter air pollution and daily mortality counts. Substantial
work has been done to study this question and to date, there appears to
be strong evidence of an association between particulate matter less
than 10 \(\mu\)g/m\(^3\) in aerodynamic diameter (PM10) and daily
mortality from all non-accidental causes \citep{samet2000fine}. For our
second example, we use the problem of studying PM10 and mortality along
with data from the National Morbidity, Mortality, and Air Pollution
Study (NMMAPS) to demonstrate how our analysis validation checks
described in Section~\ref{sec-method} can be applied. The dataset that
we use to develop our simulations and analysis checks is from New York
City, and contains daily PM10, all-cause (non-accidental) mortality, and
average temperature values from 1992--2000. We then will apply the
analysis checks to similar data from 39 other cities from the NMMAPS
study.

@@

For the purposes of this application, observing an estimated coefficient
outside of the expected interval is considered unusual and might warrant
a re-examination of the analysis process. Therefore, our unexpected
outcome in this analysis will be the binary indicator of whether the
estimated coefficient for PM10 in the GLM lies outside of the interval
\([0, 0.005]\). In addition to providing a more substantial problem for
our methods, this example also demonstrates how the procedure presented
in Section~\ref{sec-method} can be used to select cutoff values in the
analysis checks to diagnose an unexpected PM10 coefficient from the
generalized linear model.

This PM10 coefficient expectation can be framed as an analysis check
that fails, labelled as 1, if the estimate of the PM10 coefficient
(adjusted for temperature) is outside the range {[}0, 0.005{]}, and 0 if
it is within. Multiple factors can affect the estimated PM10
coefficient, such as the strength of the correlation between mortality
and PM10, or between mortality and temperature. Outliers in the three
variables can also leverage the coefficient. While these are possible
factors that could affect the analysis result, it is not clear what the
cutoff values for these checks should be to determine a failure. Here we
consider a list of checks in Table~\ref{tbl-checks} with varied cutoff
values for each:

@@

A logic regression is fitted using all variations of the checks in
Table~\ref{tbl-checks} to predict whether the PM10 coefficient is
unexpected. Given the imbalance of outcome (expected vs.~unexpected),
inverse weights proportional to the number of observations in each
outcome are applied during the logic regression fit.
Figure~\ref{fig-linear-reg-tree} shows the optimal logic regression tree
from the fitted model. Precision, recall, and independence score, along
with their harmonic and arithmetic mean are calculated in
Table~\ref{tbl-linear-reg}.

@@

\caption{\label{tbl-linear-reg}Accuracy (precision and recall) and
parsimony (independence) metrics derived from the logic regression model
(last row) and individual checks, along with harmonic and arithmetic
means.}

@@

\begin{itemize}
\tightlist
\item
  mortality-PM10 correlation less than \(-0.03\)
\item
  mortality-temperature correlation greater than \(-0.3\)
\item
  PM10 data contains outliers that are detected by the univariate
  outlier detection
\item
  mortality data contains outliers that are detected by the univariate
  outlier detection
\end{itemize}

The tree structure suggests checking mortality-PM10 correlation and a
sample size larger than 200 with an additional check of either outlier
on mortality or correlation between mortality and temperature. This
combined check rule generates a precision of 0.722 and a recall of 0.876
for predicting the unexpected PM10 coefficient.

As shown in Figure~\ref{fig-linear-reg-tree}, there is no single
analysis check in the tree that predicts an unexpected outcome. rather
at least three checks in the tree must be TRUE in order for the model to
predict an unexpected outcome. Given the high independence of the checks
(Table~\ref{tbl-linear-reg}), this suggests that unexpected results are
only likely after multiple anomalies are observed in the data.

@@

It is clear from Table~\ref{tbl-all-cities} that some of the estimated
PM10 coefficients are outside the expected range of {[}0, 0.005{]}.
Eight of the 40 cities had negative estimated coefficients and were
therefore considered unexpected by our original criterion. (It is
perhaps worth noting that none of the unexpected outcomes was in the
positive direction.) Table~\ref{tbl-accuracy} summarizes the prediction
accuracy of the logic regression model for the 40 cities. The model
correctly predicts the PM10 coefficient status of 32 cities, producing
an accuracy rate of 80\%. Among the 8 cities with an unexpected outcome
(PM10 coefficient outside {[}0, 0.005{]}), the model correctly
identifies 7 (Atlanta, Austin, Memphis, Orlando, Salt Lake City, San
Antonio, Tampa), giving a recall of 88\% (only Miami was not properly
classified by the logic regression model). These cities were flagged due
to failures in both the mortality-PM10 correlation and
mortality-temperature correlation analysis checks. Out of the 14 cities
with positive predictions from the model, 7 cities (Cincinnati,
Columbus, Dallas/Fort Worth, Pittsburgh, Riverside, San Bernardino,
Santa Ana/Anaheim) were false positives, resulting in a precision of
only 50\%. This precision value is substantially lower than what was
estimated by the simulation procedure (see Table~\ref{tbl-linear-reg})
and suggests that the simulation does not adequately capture some
features of the data generation process.

@@

With the analysis validation checks proposed in this paper, analysts may
perform these checks \emph{ex ante} to identify and address potential
issues before proceeding with the analysis or use them \emph{ex post} to
diagnose the analysis when an unexpected outcome occurs. Theses checks
act as checkpoints designed by analysts to flag potential issues,
clarify assumptions made during the analysis, and enhance transparency
and communication among stakeholders. They also support students and
junior researchers develop the skills needed to independently diagnose
issues in analyses.

@@

Visualization methods are also valuable tools for assessing data
assumptions and can potentially be formalized as analysis validation
checks. For instance, plotting a variable's distribution using a
histogram, density plot, or bee swarm plot can reveal outliers or
deviations from normality. These visualizations could be re-framed as
checks that fail when the data does not conform to the expected
distribution. However, translating visualization results into binary
checks that return (TRUE, FALSE) remains an open challenge, requiring
either manual verification or the development of automated methods to
interpret visualization outputs. An existing example of visual test is
the R package \texttt{vdiffr} \citep{vdiffr} for software unit testing.
The package saves a template plot and compares it to the current plot to
determine whether the unit tests passes or fails.

>>>>

In the study of the health effects of outdoor air pollution, one area of
interest is the association between short-term, day-to-day changes in
particulate matter air pollution and daily mortality counts. Substantial
work has been done to study this question and to date, there appears to
be strong evidence of an association between particulate matter less
than 10 \(\mu\)g/m\(^3\) in aerodynamic diameter (PM10) and daily
mortality from all non-accidental causes \citep{samet2000fine}. For our
second example, we use the problem of studying PM10 and mortality along
with data from the National Morbidity, Mortality, and Air Pollution
Study (NMMAPS) to demonstrate how our analysis validation checks
described in Section~\ref{sec-method} can be applied. In addition to
providing a more substantial problem for our methods, this example also
demonstrates how the procedure presented in Section~\ref{sec-method} can
be used to select cutoff values in the analysis checks to diagnose an
unexpected PM10 coefficient from the generalized linear model. The
dataset that we use to develop our simulations and analysis checks is
from New York City, and contains daily PM10, all-cause (non-accidental)
mortality, and average temperature values from 1992--2000. We then will
apply the analysis checks to similar data from 39 other cities from the
NMMAPS study.

@@

Multiple factors can affect the estimated PM10 coefficient, such as the
strength of the correlation between mortality and PM10, or between
mortality and temperature. Outliers in the variables can also leverage
the coefficient. While these are possible factors that could affect the
analysis result, it is not clear what the cutoff values for these checks
should be to determine a failure. Here we consider a list of checks in
Table~\ref{tbl-checks} with varied cutoff values:

@@

A logic regression is fitted using all variations of the checks in
Table~\ref{tbl-checks} to predict whether the PM10 coefficient is
unexpected. Given the imbalance of outcome (expected vs.~unexpected),
inverse weights proportional to the number of observations in each
outcome are applied during the logic regression fit.
Figure~\ref{fig-linear-reg-tree} shows the optimal logic regression tree
from the fitted model. Precision, recall, and independence score, along
with their harmonic and arithmetic mean are presented in
Table~\ref{tbl-linear-reg}.

@@

\caption{\label{tbl-linear-reg}Precision, recall, and independence
metrics derived from the logic regression model and each individual
check, along with harmonic and arithmetic means.}

@@

\begin{itemize}
\tightlist
\item
  mortality-PM10 correlation less than \(-0.03\)
\item
  mortality-temperature correlation greater than \(-0.3\)
\item
  PM10 data contains outliers detected by the univariate outlier
  detection
\item
  mortality data contains outliers detected by the univariate outlier
  detection
\end{itemize}

The fitted logic regression model is

\begin{quote}
(cor(mortality, PM10) \textless{} -0.03) AND (cor(mortality, tmp)
\textgreater{} -0.3 OR (mortality outlier AND PM10 outlier))
\end{quote}

This generates a precision of 0.722 and a recall of 0.876 for predicting
the unexpected PM10 coefficient. As shown in
Figure~\ref{fig-linear-reg-tree}, there is no single analysis check in
the tree that predicts an unexpected outcome. rather at least three
checks in the tree must be TRUE in order for the model to predict an
unexpected outcome. Given the high independence of the checks
(Table~\ref{tbl-linear-reg}), this suggests that unexpected results are
only likely after multiple anomalies are observed in the data.

@@

It is clear from Table~\ref{tbl-all-cities} that some of the estimated
PM10 coefficients are outside the expected range of {[}0, 0.005{]}.
Eight of the 40 cities had negative estimated coefficients and were
therefore considered unexpected by our original criterion. (It is
perhaps worth noting that none of the unexpected outcomes was in the
positive direction.) Table~\ref{tbl-accuracy} summarizes the prediction
accuracy of the logic regression model for the 40 cities. The model
correctly predicts the PM10 coefficient status of 32 cities, producing
an accuracy rate of 80\%. Among the 8 cities with an unexpected outcome,
the model correctly identifies 7 (Atlanta, Austin, Memphis, Orlando,
Salt Lake City, San Antonio, Tampa), giving a recall of 88\% (only Miami
was not properly classified by the logic regression model). These cities
were flagged due to failures in both the mortality-PM10 correlation and
mortality-temperature correlation analysis checks. Out of the 14 cities
with positive predictions from the model, 7 cities (Cincinnati,
Columbus, Dallas/Fort Worth, Pittsburgh, Riverside, San Bernardino,
Santa Ana/Anaheim) were false positives, resulting in a precision of
only 50\%. This precision value is substantially lower than what was
estimated by the simulation procedure (see Table~\ref{tbl-linear-reg})
and suggests that the simulation does not adequately capture some
features of the data generation process.

@@

@@

Visualization methods are also valuable tools for assessing data
assumptions and can potentially be formalized as analysis validation
checks. For instance, plotting a variable's distribution using a
histogram, density plot, or bee swarm plot can reveal outliers or
deviations from normality. These visualizations could be re-framed as
analysis checks that fail when the data does not conform to the visual
expectation. However, translating visualization results into binary
checks remains an open challenge, requiring either manual verification
or the development of automated methods to interpret visualization
outputs. An existing example of visual test is the R package
\texttt{vdiffr} \citep{vdiffr} for graphic software unit testing. The
package saves a template plot and compares it to the current plot to
determine whether the unit tests passes or fails.

### 2024-12-24 index.qmd
section 5 and 6

In the study of the health effects of outdoor air pollution, one area of interest is the association between short-term, day-to-day changes in particulate matter air pollution and daily mortality counts. Substantial work has been done to study this question and to date, there appears to be strong evidence of an association between particulate matter less than 10 $\mu$g/m$^3$ in aerodynamic diameter (PM10) and daily mortality from all non-accidental causes [@samet2000fine]. For our second example, we use the problem of studying PM10 and mortality along with data from the National Morbidity, Mortality, and Air Pollution Study (NMMAPS) to demonstrate how our analysis validation checks described in @sec-method can be applied. The dataset that we use to develop our simulations and analysis checks is from New York City, and contains daily PM10, all-cause (non-accidental) mortality, and average temperature values from 1992--2000. We then will apply the analysis checks to similar data from 39 other cities from the NMMAPS study.

@@

For the purposes of this application, observing an estimated coefficient outside of the expected interval is considered unusual and might warrant a re-examination of the analysis process. Therefore, our unexpected outcome in this analysis will be the binary indicator of whether the estimated coefficient for PM10 in the GLM lies outside of the interval $[0, 0.005]$. In addition to providing a more substantial problem for our methods, this example also demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the analysis checks to diagnose an unexpected PM10 coefficient from the generalized linear model.

This PM10 coefficient expectation can be framed as an analysis check that fails, labelled as 1, if the estimate of the PM10 coefficient (adjusted for temperature) is outside the range [0, 0.005], and 0 if it is within. Multiple factors can affect the estimated PM10 coefficient, such as the strength of the correlation between mortality and PM10, or between mortality and temperature. Outliers in the three variables can also leverage the coefficient. While these are possible factors that could affect the analysis result, it is not clear what the cutoff values for these checks should be to determine a failure. Here we consider a list of checks in @tbl-checks with varied cutoff values for each:

@@

A logic regression is fitted using all variations of the checks in @tbl-checks to predict whether the PM10 coefficient is unexpected. Given the imbalance of outcome (expected vs. unexpected), inverse weights proportional to the number of observations in each outcome are applied during the logic regression fit. @fig-linear-reg-tree shows the optimal logic regression tree from the fitted model. Precision, recall, and independence score, along with their harmonic and arithmetic mean are calculated in @tbl-linear-reg.

@@

```{r}
#| fig.width: 12
#| fig.height: 4
fit <- pm10_df |> fit_logic_reg(unexpect, `cor(m, PM10) < -0.05`:mortality_outlier,
                                seed = 7, nleaves = 4, wgt = wgt_vec)
plot_logicreg(fit)
```

@@

```{r}
#| label: tbl-linear-reg
#| tbl-cap: "Accuracy (precision and recall) and parsimony (independence) metrics derived from the logic regression model (last row) and individual checks, along with harmonic and arithmetic means."
df <- list(
  tibble(.fitted = pm10_df$`cor(m, PM10) < -0.03`, 
         unexpect = pm10_df$unexpect),
  tibble(.fitted = pm10_df$`cor(m, tmp) > -0.3`, 
         unexpect = pm10_df$unexpect),
  tibble(.fitted = pm10_df$mortality_outlier, 
         unexpect = pm10_df$unexpect),
  tibble(.fitted = pm10_df$PM10_outlier, 
         unexpect = pm10_df$unexpect),
  augment(fit)) |>
  map_dfr(~.x |> calc_miscla_rate(.fitted, unexpect) |> 
            mutate(overlapping = 0, independence = 1) |> 
            calc_metrics(metrics = c("harmonic", "arithmetic")), .id = "id")

@@

-   mortality-PM10 correlation less than $-0.03$
-   mortality-temperature correlation greater than $-0.3$
-   PM10 data contains outliers that are detected by the univariate outlier detection
-   mortality data contains outliers that are detected by the univariate outlier detection

The tree structure suggests checking mortality-PM10 correlation and a sample size larger than 200 with an additional check of either outlier on mortality or correlation between mortality and temperature. This combined check rule generates a precision of `r round(as.numeric(df[5, "precision"]), 3)` and a recall of `r round(as.numeric(df[5, "recall"]), 3)` for predicting the unexpected PM10 coefficient.

As shown in @fig-linear-reg-tree, there is no single analysis check in the tree that predicts an unexpected outcome. rather at least three checks in the tree must be TRUE in order for the model to predict an unexpected outcome. Given the high independence of the checks (@tbl-linear-reg), this suggests that unexpected results are only likely after multiple anomalies are observed in the data.

@@

It is clear from @tbl-all-cities that some of the estimated PM10 coefficients are outside the expected range of [0, 0.005]. Eight of the 40 cities had negative estimated coefficients and were therefore considered unexpected by our original criterion. (It is perhaps worth noting that none of the unexpected outcomes was in the positive direction.) @tbl-accuracy summarizes the prediction accuracy of the logic regression model for the 40 cities. The model correctly predicts the PM10 coefficient status of `r accurate <- df |> filter(unexpect == .pred) |> nrow(); accurate` cities, producing an accuracy rate of `r label_percent()(accurate/nrow(df))`. Among the `r df |> filter(unexpect) |> nrow()` cities with an unexpected outcome (PM10 coefficient outside [0, 0.005]), the model correctly identifies `r df |> filter(unexpect, .pred) |> nrow()` (`r cities_accurate`), giving a recall of `r recall <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(recall)); label_percent()(recall)` (only `r cities_missed` was not properly classified by the logic regression model). These cities were flagged due to failures in both the mortality-PM10 correlation and mortality-temperature correlation analysis checks. Out of the `r df |> filter(.pred) |> nrow()` cities with positive predictions from the model, `r df |> filter(unexpect != .pred, !unexpect) |> nrow()` cities (`r df |> filter(unexpect != .pred, !unexpect) |> pull(cityname)`) were false positives, resulting in a precision of only `r pres <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(precision)); label_percent()(pres)`. This precision value is substantially lower than what was estimated by the simulation procedure (see @tbl-linear-reg) and suggests that the simulation does not adequately capture some features of the data generation process.

@@

With the analysis validation checks proposed in this paper, analysts may perform these checks *ex ante* to identify and address potential issues before proceeding with the analysis or use them *ex post* to diagnose the analysis when an unexpected outcome occurs. Theses checks act as checkpoints designed by analysts to flag potential issues, clarify assumptions made during the analysis, and enhance transparency and communication among stakeholders. They also support students and junior researchers develop the skills needed to independently diagnose issues in analyses.

@@

Visualization methods are also valuable tools for assessing data assumptions and can potentially be formalized as analysis validation checks. For instance, plotting a variable's distribution using a histogram, density plot, or bee swarm plot can reveal outliers or deviations from normality. These visualizations could be re-framed as checks that fail when the data does not conform to the expected distribution. However, translating visualization results into binary checks that return (TRUE, FALSE) remains an open challenge, requiring either manual verification or the development of automated methods to interpret visualization outputs. An existing example of visual test is the R package `vdiffr` [@vdiffr] for software unit testing. The package saves a template plot and compares it to the current plot to determine whether the unit tests passes or fails.

>>>>

In the study of the health effects of outdoor air pollution, one area of interest is the association between short-term, day-to-day changes in particulate matter air pollution and daily mortality counts. Substantial work has been done to study this question and to date, there appears to be strong evidence of an association between particulate matter less than 10 $\mu$g/m$^3$ in aerodynamic diameter (PM10) and daily mortality from all non-accidental causes [@samet2000fine]. For our second example, we use the problem of studying PM10 and mortality along with data from the National Morbidity, Mortality, and Air Pollution Study (NMMAPS) to demonstrate how our analysis validation checks described in @sec-method can be applied. In addition to providing a more substantial problem for our methods, this example also demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the analysis checks to diagnose an unexpected PM10 coefficient from the generalized linear model. The dataset that we use to develop our simulations and analysis checks is from New York City, and contains daily PM10, all-cause (non-accidental) mortality, and average temperature values from 1992--2000. We then will apply the analysis checks to similar data from 39 other cities from the NMMAPS study.

@@

Multiple factors can affect the estimated PM10 coefficient, such as the strength of the correlation between mortality and PM10, or between mortality and temperature. Outliers in the variables can also leverage the coefficient. While these are possible factors that could affect the analysis result, it is not clear what the cutoff values for these checks should be to determine a failure. Here we consider a list of checks in @tbl-checks with varied cutoff values:

@@

A logic regression is fitted using all variations of the checks in @tbl-checks to predict whether the PM10 coefficient is unexpected. Given the imbalance of outcome (expected vs. unexpected), inverse weights proportional to the number of observations in each outcome are applied during the logic regression fit. @fig-linear-reg-tree shows the optimal logic regression tree from the fitted model. Precision, recall, and independence score, along with their harmonic and arithmetic mean are presented in @tbl-linear-reg.

@@

```{r}
#| fig.width: 12
#| fig.height: 6
fit <- pm10_df |> fit_logic_reg(unexpect, `cor(m, PM10) < -0.05`:mortality_outlier,
                                seed = 7, nleaves = 4, wgt = wgt_vec)
plot_logicreg(fit)
```

@@

```{r}
#| label: tbl-linear-reg
#| tbl-cap: "Precision, recall, and independence metrics derived from the logic regression model and each individual check, along with harmonic and arithmetic means."
df <- list(
  tibble(.fitted = pm10_df$`cor(m, PM10) < -0.03`, 
         unexpect = pm10_df$unexpect),
  tibble(.fitted = pm10_df$`cor(m, tmp) > -0.3`, 
         unexpect = pm10_df$unexpect),
  tibble(.fitted = pm10_df$mortality_outlier, 
         unexpect = pm10_df$unexpect),
  tibble(.fitted = pm10_df$PM10_outlier, 
         unexpect = pm10_df$unexpect),
  augment(fit)) |>
  map_dfr(~.x |> calc_miscla_rate(.fitted, unexpect) |> 
            mutate(overlapping = 0, independence = 1) |> 
            calc_metrics(metrics = c("harmonic", "arithmetic")), .id = "id")

@@

-   mortality-PM10 correlation less than $-0.03$
-   mortality-temperature correlation greater than $-0.3$
-   PM10 data contains outliers detected by the univariate outlier detection
-   mortality data contains outliers detected by the univariate outlier detection

The fitted logic regression model is

  > (cor(mortality, PM10) < -0.03) AND (cor(mortality, tmp) > -0.3 OR (mortality outlier AND PM10 outlier))

This generates a precision of `r round(as.numeric(df[5, "precision"]), 3)` and a recall of `r round(as.numeric(df[5, "recall"]), 3)` for predicting the unexpected PM10 coefficient. As shown in @fig-linear-reg-tree, there is no single analysis check in the tree that predicts an unexpected outcome. rather at least three checks in the tree must be TRUE in order for the model to predict an unexpected outcome. Given the high independence of the checks (@tbl-linear-reg), this suggests that unexpected results are only likely after multiple anomalies are observed in the data.

@@

It is clear from @tbl-all-cities that some of the estimated PM10 coefficients are outside the expected range of [0, 0.005]. Eight of the 40 cities had negative estimated coefficients and were therefore considered unexpected by our original criterion. (It is perhaps worth noting that none of the unexpected outcomes was in the positive direction.) @tbl-accuracy summarizes the prediction accuracy of the logic regression model for the 40 cities. The model correctly predicts the PM10 coefficient status of `r accurate <- df |> filter(unexpect == .pred) |> nrow(); accurate` cities, producing an accuracy rate of `r label_percent()(accurate/nrow(df))`. Among the `r df |> filter(unexpect) |> nrow()` cities with an unexpected outcome, the model correctly identifies `r df |> filter(unexpect, .pred) |> nrow()` (`r cities_accurate`), giving a recall of `r recall <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(recall)); label_percent()(recall)` (only `r cities_missed` was not properly classified by the logic regression model). These cities were flagged due to failures in both the mortality-PM10 correlation and mortality-temperature correlation analysis checks. Out of the `r df |> filter(.pred) |> nrow()` cities with positive predictions from the model, `r df |> filter(unexpect != .pred, !unexpect) |> nrow()` cities (`r df |> filter(unexpect != .pred, !unexpect) |> pull(cityname)`) were false positives, resulting in a precision of only `r pres <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(precision)); label_percent()(pres)`. This precision value is substantially lower than what was estimated by the simulation procedure (see @tbl-linear-reg) and suggests that the simulation does not adequately capture some features of the data generation process.

@@

@@

Visualization methods are also valuable tools for assessing data assumptions and can potentially be formalized as analysis validation checks. For instance, plotting a variable's distribution using a histogram, density plot, or bee swarm plot can reveal outliers or deviations from normality. These visualizations could be re-framed as analysis checks that fail when the data does not conform to the visual expectation. However, translating visualization results into binary checks remains an open challenge, requiring either manual verification or the development of automated methods to interpret visualization outputs. An existing example of visual test is the R package `vdiffr` [@vdiffr] for graphic software unit testing. The package saves a template plot and compares it to the current plot to determine whether the unit tests passes or fails.
